---
title: Anonymize Individuals using digest()
author: Xing
date: '2017-12-14'
slug: anonymize-using-digest
categories: []
tags:
  - Data Science
  - Experiments
  - R
  - Privacy
header:
  caption: ''
  image: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

When requesting individual level data from others (a company or a government agency), we usually need to properly anomymize the individuals to protect their privacy. The following is an example:

```{r Data}
(Data = data.frame(Name = c("John Smith", "Jenny Ford","Vivian Lee"), 
                  Secret = c("Hate dog","Afraid of ghost","A bathroom dancer")))
```

One simple way is we can just drop the **Name**, and only keep the **Secret** since we are more interested in their secrets. But what if we need to update individuals' information periodically? We need a unique identifier for each individual so that it can link to his/her new information. The tricky part is that we cannot use their name as the identifier. We have to create a "cyphertext"" that only the data provider can interpret but we cannot. 

In many cases, the data providers do not have the capacity to create a cyphertext in a systematic way. They either 1) do not provide the data due to privacy concerns, or 2) the data does not an identifier we can use in the future. Here I use R's `digest()` package to show how to anomymize individuals and generate an identifier using `sha1` algorithm. `sha1` (Secure Hash Algorithm 1) is a cryptographic hash function which takes an input and produces a 40 digits long hash value known as a message digest. The user needs to have the random seed to decrypt the hash value and recover the original value. We can just pass the code to the data provider, and ask them to run it (R is free!). What they need to remember is jus the random seed.


```{r digest}
library(digest)
# Suppose variable Name is the variable to be hashed
# But we have to make the string to be hashed as character
Data$CharName <- as.character(Data$Name)
# sha1() function only work on one string each time
# Hence we first vectorize this function so it can work on a column
vSha1 <- Vectorize(sha1)

#########################################################################
## Set the random see for the hash function                            ##
## Procedurally, if you want to make sure that youself cannot recover  ##
## the original name, you let them to set the seed and run the code.   ##
SeedForDigest = 200 
#########################################################################

## Hash it!
# To avoid serializing the vector, we set serialize=FALSE

Data$HashedName = vSha1(Data$CharName, algo=c("sha1"), serialize=FALSE, file=FALSE, 
                     length=Inf, skip="auto", ascii=FALSE, raw=FALSE, seed=SeedForDigest, 
                     errormode=c("warn"))
### Drop the original name
Data$Name = NULL
Data$CharName = NULL
Data
```