---
title: Personalized Data Summary Function Using "data.table"
author: Xing
date: '2018-01-19'
slug: personalized-data-summary-function-using-data-table
categories: []
tags:
  - R
  - Data
  - Statistics
output:
 blogdown::html_page:
    toc: false  
---

One function I miss about *Stata* is its *tabstat*. By using just one line code, it can produce very useful summary statistics such as `mean`, and `standard error` by groups by conditions. R has its own built-in summary function -- `summary()`, too, but in most cases in my research, I found the summaries produced is barely useful. Consider the following pseudo-data:

```{r Gendata}
library(data.table)
set.seed(10)
N = 120
DT = data.table(x = rnorm(N,1), y = rnorm(N,2),
                category = sample(letters[1:3], N, replace = T))
DT[1:10]
```

If we summarize the variable `x` using `summary()`, it gives:

```{r Rsummary}
summary(DT$x)
```

In most the case, I want to have a sense of the dispersion of the mean, number of non-mising observations. More importantly, I want to have a data table from which I can generate a barchart, which is very common in analyzing experimental data. Since I almost need this types of summary function for all of my on-going project, why not make a personalized one for myself and potential other users? Here it is. 

```{r SingleV}
SumFunOne = function(Data, Var, Group, StatList){
  arguments <- as.list(match.call())
  x = eval(arguments$Var, Data)
  category = eval(arguments$Group, Data)
  keep = StatList
  result = Data[,  .(Mean = mean(x, na.rm=TRUE), 
                     N = sum(!is.na(x)), 
                     SE = sd(x, na.rm=TRUE)/sqrt(sum(!is.na(x))),
                     median = median(x),
                     max = max(x),
                     min = min(x),
                     Missing = sum(is.na(x))),
                     by = .(category)][,..keep][order(category)]
  return(result)
}
(Data.Summary = SumFunOne(DT, x, category, c("category","Mean","N","SE")))
```

The function `SumFunOne()` has 4 inputs: `Data` (should be in `data.table`), `Var` -- variable to be summarized, `Group` -- the group variable I want to condition on, and `StatList` -- the statistics I want to show. We can also subset the data using `DT[y>0]` in the input. Given the results, I can easily draw a barchart with standard errors and number of observations in `ggplot2`:


```{r ShowBarChart, fig.width=8, fig.height=6,echo = FALSE}
library(data.table)
library(ggplot2)
ggplot(Data.Summary, aes(x = category, y = Mean)) +
  geom_bar(position="dodge", stat = "identity",fill="black") +
  geom_errorbar(aes(ymin=Mean - SE, ymax=Mean + SE), 
                position=position_dodge(0.9), width=.2,col="grey45")+
  geom_text(aes(label= N), vjust=1.5, colour="white",
            position=position_dodge(.9), size=8) +
  theme(legend.position="bottom", legend.text=element_text(size=24), 
        legend.title = element_text(size = "24"), 
        plot.title = element_text(size = rel(4), hjust = 0.5),
        axis.title.y = element_text(size = rel(3)), 
        axis.title.x = element_text(size = rel(3)),
        axis.text.x = element_text(angle=0, vjust=1,hjust=1, size=30), 
        axis.text.y = element_text(angle=0, vjust=1, size=30),
        strip.text = element_text(size=30),
        axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        strip.background = element_blank())+ 
  labs(title = "", 
       x = "Category", y = "Mean") 
```

Well, what if I want to summarize multiple variables? The goal is to have a counterpart of Stata's *tabstat* in R, isn't it? It is straightforward, too. We just need to use the powerful `.SD` in `data.table` to apply the summary function to multiple variables. But we define the summary function first outside of the data.table. For simplicity, I only show three statistics: `Mean`, `N` and `SE`. `varList` is the variable list we want to summrize.

```{r multiVariable}

SumFunMult = function(Data, varList, Group){
  arguments <- as.list(match.call())
  category = eval(arguments$Group, Data)
  my.summary <- function(x){
    c(Mean = mean(x, na.rm=TRUE), 
      N = sum(!is.na(x)), 
      SE = sd(x, na.rm=TRUE)/sqrt(sum(!is.na(x))))
  }
  result = Data[, lapply(.SD, my.summary), by=.(category), .SDcols= varList]
  Stats = rep(c("Mean","N","SE"),length(unique(category)))
  summary = cbind(Stats,result)
  return(summary)
}

SumFunMult(DT, c("x","y"),category)

```

Next task is the develop it into a package so that I can easily call the function to summarize and visualize the summary statistics...

